RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-11-04 manuscripts/manuscript.Rmd

Though certain functions allow R users to switch between these formats
within R, these formats frequently contain more information than their
NEXUS file counterparts, and users lack a suitable alternative for
archiving and sharing phylogenetic data outside of the R ecosystem. 
Consequently, information added to phylogenies by these R programs cannot 
easily be shared, archived, and made available to other software;
but inherits the limitations of the NEXUS format.

@@

<!-- FIXME something about how we will increasingly want to work with (semantic) metadata
and how RNeXML provides a solid platform from which to do so.  
-->

@@

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in R formats
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

@@

A NeXML document can be thought of as having multiple components: (1) phylogeny
data (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
`RNeXML` seeks to provide a seamless way to convert these to and from their 
corresponding R object representations, such as `ape::phylo` phylogenies, character
matrices (`data.frames`) and metadata lists (or for advanced users,
`XMLInternalDocument` and `rrdf` triplestores). Metadata can provide
information about the study in which the phylogeny was generated or
applied (such as authors, citations to publications, timestamps, and even
descriptions of methods), but can also be used to annotate individual
data components.  We illustrate how metadata annotations can be added to
the operational taxonomic units of the phylogeny and can be used to extend
the data representation to new uses such as stochastic character maps.

@@

```{r}
library("RNeXML")
f <- system.file("examples", "trees.xml", package="RNeXML")
nex <- nexml_read(f)
```

The resulting `nex` object is an R object representation of the NeXML. This is 
a native R object, of the `nexml` class, for representing phylogenetic trees (as well as 
character traits, and metadata), analogous to the `ape::phylo` object from the 
popular `ape` R package [@Paradis_2004]. While R functions and packages 
could be written to use this `nexml` object directly, most users will prefer to use
already existing methods provided by packages they know. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
case we will get back a `multiPhylo` object, since the file contains two phylogenies.

@@

NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees()` function attempts to return the most sensible 
R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.
While most users will find this behavior convenient, there are times
(such as looping over many NeXML files) when having a variable return
type is undesirable.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in 
[TreeBASE](http://treebase.org/treebase-web/home.html) <!-- first mention -->
<!-- of TreeBase, include URL? or introduce it in intro? --> through the
`nexml` format.  `RNeXML` can read directly from a URL:

```{r eval=FALSE}
nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
```

Previously, this was possible using the `treebase` package [@Boettiger_2012]
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard [@Vos_2012], 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NeXML standard, every valid NeXML document on TreeBASE can now successfully
be parsed to R objects.

@@

Like NEXUS files, NeXML documents can also contain character data.  While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence (or morphological) data 
used to infer the phylogenetic tree, we find this ideal for character data used in
comparative methods popular in R. Methods to extract character data
work much like those for the phylogenetic data. Here we read in an example
NeXML file that contains both a phylogeny and character data. 
`get_characters()` returns a `data.frame` with columns as characters and rows as
taxa:

```{r }
comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
nex <- nexml_read(comp_analysis)
get_characters(nex)
```

As this example shows, continuous and discrete characters are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. The `get_characters_list()` function will consistently receive
a list of data frames corresponding to each character matrix block of
the NeXML.

@@

The first argument to `nexml_write()` can be a `ape::phylo` phylogeny,
`multiPhylo` list of phylogenies, or a `nexml` object -- anything that
can be coerced into the `nexml` class.  As we have seen, NeXML files
can contain multiple phylogenetic trees, as well as character matrices.
We can add both trees and character data using `nexml_write()`,

@@

NeXML files can also be built up bit by bit using `add_characters()`,
`add_trees()`, and `add_meta()`.  Each of these functions can add data to an
existing `nexml` object or create a new nexml object if none is provided.
Here we first create a `nexml` object containing the phylogeny data,
and then add the character data to it:

@@

This allows us to include multiple trees and character matrices to a single NeXML file.

@@

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compatibility with any existing NeXML parser. NeXML uses the
"Resource Description Framework in Annotations", or RDFa,
as described in @Vos_2012 or @W3C_2014. 
These annotations can be used to add 
information to any particular element of the data, (a tree, node, edge,
or taxonomic unit; a character matrix or specific trait, or even another 
metadata annotation or the NeXML data file as a whole). `RNeXML` provides
a range of tools for accessing metadata embedded in a NeXML file.

@@

Researchers cannot take advantage of NeXML metadata manipulation if
there is no metadata there.  The burden of adding good metadata remains a
significant barrier to increased sharing and reuse of scientific research
[@Tenopir_2011].
`RNeXML` seeks to reduce the burden of adding
high quality, machine-readable metadata to NeXML files. This occurs on
several levels:

@@

`RNeXML` automatically adds basic metadata by default such as a time-stamp
of when the tree was produced and an open data license (CC0, compatible
with deposition on Dryad data repository), which can be configured
as needed.

To adjust these and other common metadata fields that should 
be part of any data deposition, `RNeXML` provides the `add_basic_meta()`
function.  This function does not assume any familiarity with the concepts
of namespaces or other more advanced metadata tools discussed later.  The
`add_basic_meta()` function takes an existing `nexml` object (or will create a new one
if none is provided, like the other `add_` functions), and any of the following
fields: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Additional fields may be added to subsequent versions, see
`?add_basic_metadata` for the most up-to-date information.  Here we create
a new `nexml` object with the basic metadata for the `bird.orders` data
provided in the `ape` package:

```{r}
data(bird.orders)
birds <- add_trees(bird.orders)
birds <- add_basic_meta(birds,
  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",

@@

Citations can also be added using R's `bibentry` type, which allows
a user to add citations to R packages,

@@

or to papers by simply using the object's DOI (using @Boettiger_2014).
For instance to add the citation information of the paper that generated the
`geospiza` phylogeny included in the `geiger` package:

@@

`RNeXML` provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.
One common use case is that of taxonomic identifiers.  Misspellings or
inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration in reusing phylogenetic trees, especially when
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come from
other places that use different taxonomies. Hence, not just misspellings,
but also synonyms, homonyms and incompatible taxon concepts might plague
the end user. By associating the taxonomic labels against a specific,
centralized database we can catch some of these inconsistencies and
indicate more precisely which classification we are appealing to in the
event of conflicting taxonomic definitions.  In this example, we indicate
that we intend the bird orders listed in the tree to correspond to the
NCBI definitions.

>>>>

Though certain functions allow R users to switch between these formats
within R, these formats frequently contain more information than their
NEXUS file counterparts,
<!-- such as what? we haven't talked about metadata yet. HL -->
and users lack a suitable alternative for
archiving and sharing phylogenetic data outside of the R ecosystem. 
Consequently, information added to phylogenies by these R programs cannot 
easily be shared, archived, and made available to other software;
<!-- such as what? should walk through an example here. HL -->
but inherits the limitations of the NEXUS format.
<!-- how can information inherit the limitations of a serialization -->
<!-- format if it's added to data read from that format? HL -->

@@

<!-- FIXME something about how we will increasingly want to work with (semantic) metadata
and how RNeXML provides a solid platform from which to do so.
-->
<!-- I think this is in fact more important than the paragraph above. I.e., -->
<!-- whether exchanging basic comparative data needs another format is -->
<!-- debatable, but that metadata that can hardly be expressed in -->
<!-- current standards does is a clear point. HL -->

@@

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses metadata to
extend the NeXML standard to describe stochastic character mapping, one
of the kinds of data annotation that is currently available in R
formats
<!-- R packages, not formats, right? HL -->
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

@@

A NeXML document can be thought of as having the following components: (1) phylogeny
data, (2) character-trait data, (3) operational taxonomic units, and (4) metadata. 
`RNeXML` seeks to provide a seamless way to convert these to and from their 
corresponding R object representations, in particular `ape::phylo`,
`data.frame`, and lists for phylogenies, character
matrices, and metadata, respectively.
<!-- I feel the following is a little early - it will likely be over -->
<!-- most most readers' heads. HL
(or for advanced users,
`XMLInternalDocument` and `rrdf` triplestores).
-->
Metadata can provide bibliographic and provenance
information about the study as part of which the phylogeny was generated or
applied (such as authors, citations to publications, timestamps, and even
descriptions of methods). They can also be used to attach annotations to individual
data elements.  We illustrate how metadata annotations can be added to
the operational taxonomic units of the phylogeny and can be used to extend
the data representation to new uses such as stochastic character maps.
<!-- sounds awkward - need to smoothen the language. HL -->

@@

```{r}
library("RNeXML")
f <- system.file("examples", "trees.xml", package="RNeXML")
nex <- nexml_read(f)
```
<!-- The system.file() command is not pretty. Can we not just read in -->
<!-- the file directly? HL -->
The resulting `nex` object is of type `nexml` and is a native R object representation of the NeXML
file. <!-- This seems redundant to me to what has already been -->
<!-- said. HL
Objects of type `nexml` represent phylogenetic trees as well as 
character traits, and metadata, analogous to the `ape::phylo` object from the 
popular `ape` R package [@Paradis_2004]. --> Although new custom R functions and packages 
could be written for using `nexml` objects, for most users it will be
more convienent to use methods of the many already existing R packages for 
phylogenetics. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
case we will get back a `multiPhylo` object, since the file contains two phylogenies.

@@

NeXML can not only contain multiple phylogenetic trees in a single document,
but can also group sets of trees into separate blocks (e.g. as resulting from
different workflows). The `get_trees()` function attempts to return the most sensible 
R object for NeXML: if only a single tree is available, it will return an `ape::phylo` 
object; if multiple trees are available in a single `trees` block, it will return
an `ape::multiPhylo` object (a list of `ape::phylo` objects).  If multiple
blocks are available, it will return a list of `multiPhylo` blocks.
<!-- having a function's return type being more or less unpredictable -->
<!-- sounds like a bad design pattern. Is there at least precedence for -->
<!-- this practice in other R packages? HL --> 
While most users will find this behavior convenient, there are times
(such as looping over many NeXML files) when having a variable return
type is undesirable.  To always get back a list of `multiPhylo` objects,
even when only one tree is available, use `get_trees_list()`.

The ability to read in phylogenetic trees in the NeXML format opens up a
wide and rapidly growing array of resources to the R user.  For instance,
a user can access all the phylogenies available in 
[TreeBASE](http://treebase.org/) <!-- first mention -->
<!-- of TreeBase, include URL? or introduce it in intro? --> through the
`nexml` format.  `RNeXML` can read directly from a URL:

```{r eval=FALSE}
nex <- nexml_read("https://raw.github.com/rvosa/supertreebase/master/data/treebase/S100.xml")
```
<!-- the example makes me wonder how nexml_read() distinguishes -->
<!-- between a URL and a file name as its argument. HL -->
Reading phylogenies from TreeBASE into R was already possible using
the R package `treebase` [@Boettiger_2012]. However, because it relied
on the NEXUS parser provided in the `ape` package, and because of
incompatible flavors of the NEXUS standard [@Vos_2012], 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Owing to the syntactically strict
definition of the
NeXML standard, the RNeXML package can successfully read every valid
NeXML document from TreeBASE. <!-- are there invalid ones as well that -->
<!-- it can't read? the sentence as is suggests so, and so is a little -->
<!-- awkward (it leaves open the possibility that the NEXUS files ape -->
<!-- can't read are now invalid NeXML files. Which is probably not -->
<!-- what you mean to suggest. HL -->

@@

Like NEXUS files, NeXML documents can in addition to trees also
contain character data in the form of matrices.  Character data can be
of molecular (molecular sequence alignments), discrete (most
morphological character data), or continuous type (many trait data). 
Often character data, especially molecular and morphological, is used
for phylogenetic tree reconstruction, but they are also a common input
for the many comparative phylogenetic analysis methods available in R. Methods to extract character data
work much like those for extracting the phylogenetic trees. Here we read in an example
NeXML file that contains both a phylogeny and character data. The
function
`get_characters()` can then be used to obtain a `data.frame` with columns as characters and rows as
taxa:

```{r }
comp_analysis <- system.file("examples", "comp_analysis.xml", package="RNeXML")
nex <- nexml_read(comp_analysis)
get_characters(nex)
```
<!-- same as above - why can this not be one call? HL -->
Continuous and discrete character data are represented as separate
blocks in NeXML files, but will be combined as separate columns of a
single data frame by `get_characters()` if they correspond to the same
taxa. If this behavior is undesired, the function
`get_characters_list()` can be used to obtain
a list of data frames, where list elements correspond to the character matrix blocks of
the NeXML file.

@@

The first argument to `nexml_write()` is a `nexml` object, or any
object that can be coerced to it, such as a `ape::phylo` phylogeny,
or a `multiPhylo` list of phylogenies. The `nexml_write()` function
also allows to add character data:

@@

Instead of packaging trees and characters at the time of writing a
NeXML file, the document can also be built up step by step using the
functions `add_characters()`,
`add_trees()`, and `add_meta()`.  Each of these functions can add data to an
existing `nexml` object, or create a new nexml object if none is provided.
Here we first create a `nexml` object containing the phylogeny data,
and then add the character data to it:

@@

This allows us to include multiple trees and character matrices to a
single NeXML file.
<!-- perhaps give a motivating use case for why one might -->
<!-- want to do this? HL -->

@@

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compatibility with any existing NeXML parser. NeXML uses the
"Resource Description Framework in Annotations" (RDFa),
as described in @Vos_2012 or @W3C_2014. 
These annotations can be used to add 
information to any particular element of the data, (a tree, node, edge,
or taxonomic unit; a character matrix or specific trait, or even another 
metadata annotation or the NeXML data file as a whole). `RNeXML` provides
a range of tools for accessing metadata embedded in a NeXML file.

@@

The burden typically involved with documenting and sharing good
metadata alongside the data remains a
significant barrier to increased sharing and reuse of scientific research
[@Tenopir_2011]. To promote sharing of well-documented phylogenetic data,
`RNeXML` aims to provide a simple interface to adding well-defined machine-readable metadata to NeXML files. This occurs on
several levels:

@@

`RNeXML` automatically adds basic metadata by default
<!-- what does "by default" mean here? It doesn't seem like calling -->
<!-- any RNeXML function will add such metadata if it's not there -->
<!-- already? HL -->
such as a time-stamp
of when a tree was produced, and an open data license. The latter by
default is the Creative Commons Zero (CC0) public domain waiver, which
is compatible
with deposition in the Dryad data repository, but can be configured
as a different license if needed.

To adjust these and other metadata fields commonly recommended to be
included with publicly archived data, `RNeXML` provides the `add_basic_meta()`
function.  This function hides the complexities of anchoring metadata
attribute names in common vocabularies from the user.  The
`add_basic_meta()` function takes as input an existing `nexml` object
(like the other `add_` functions, if none is provided it will create one), and at the time
of this writing any of the following
parameters: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Other metadata elements and corresponding parameters may
be added in the future.

Here we create a `nexml` object for the `bird.orders` data
provided in the `ape` package, and then add basic metadata:
<!-- I notice that other add_* methods take the nexml object as the -->
<!-- last argument, not the first as seems to be the case here. This -->
<!-- is inconsistent and should be fixed. HL -->
```{r}
data(bird.orders)
birds <- add_trees(bird.orders)
birds <- add_basic_meta(birds,
  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",

@@

Instead of a literal string, citations can also be provided in R's
`bibentry` type, which is the one in which R package citations are obtained:

@@

Alternatively, a citation to a published paper with a Digital
Object Identifier (DOI) can be provided in the form of the DOI,
which the package `knitcitations` [@Boettiger_2014] can turn into a
formatted citation using. As an example, to add the citation information of the paper that generated the
`geospiza` phylogeny included in the `geiger` package:

@@

`RNeXML` provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.
One common use case is that of taxonomic identifiers.  Misspellings or
inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration when reusing phylogenetic trees, especially when
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come from
other places that use different taxonomies. Hence, not just misspellings,
but also synonyms, homonyms and incompatible taxon concepts might plague
the end user. By associating the taxonomic labels against a specific,
centralized database we can catch some of these inconsistencies and
indicate more precisely which classification we are appealing to in the
event of conflicting taxonomic definitions.  In this example, we indicate
that we intend the bird orders listed in the tree to correspond to the
NCBI definitions.
